---
title: "basic_LMM"
output: html_document
date: "2025-12-01"
---


This is a R notebook to analyze the ratings of architectural images using linear mixed models.
Here we computed the mean ratings acorss five dimensions as the dependant variable.


# Data preparation

```{r load libraries, include=FALSE}
library(tidyverse)
# load the libraries for LMM
library(lme4)
library(lmerTest)
library(car)
library(emmeans)
```


```{r preprocess, include=FALSE}
# Read the two CSV files
# Set show_col_types = FALSE to keep console clean.
df   <- readr::read_csv("Data_exp1.csv",  show_col_types = FALSE)


# Set Plus as the reference level in Category
df$Category <- relevel(factor(df$Category), ref = "Combined+")
```


# Modelling
## Find the best model, use the manual, hypothesis-driven way
```{r}
m1 <- lmer(mean_rating ~ Category + (1 | Participant) + (1 | ImageNumber), data=df, REML=FALSE)
m2 <- lmer(mean_rating ~ Category + (1 | Participant) + (1 | ImageNumber) +      (0 + Category || Participant), data=df, REML=FALSE, control=lmerControl(optimizer="bobyqa", optCtrl=list(maxfun=2e5)))
anova(m1, m2)
```

```{r}
m3 <- lmer(mean_rating ~ Category + (1 + Category | Participant) + (1 | ImageNumber), data=df, REML=FALSE)
anova(m2, m3)

```
```{r}
AIC(m2, m3)
BIC(m2, m3)
anova(m1, m3)
```
The model m3, which included random slopes for Category on Participant with correlated slopes and intercepts, was selected as the optimal model. Although the likelihood ratio test between m2 and m3 was not significant, m3 had lower AIC and BIC values, indicating a better balance between model fit and complexity. Additionally, m3 did not exhibit convergence issues, unlike m2 and more complex models. Therefore, we concluded that m3 was the most appropriate model for our analysis. Further attempts to fit more complex models failed to converge, indicating that m3 was the most complex structure supported by the data.

## Check the output

```{r model output}
# check the result of best model
best_model <- m3
summary(best_model)
Anova(best_model, type = 'III')
# Save the model object to a file
saveRDS(best_model, file = "best_model.rds")
```

A linear mixed-effects model was fitted to examine the effect of Category on mean ratings. The model included Category as a fixed effect, random intercepts for Participant and ImageNumber, and random slopes of Category by Participant. The model revealed a significant overall effect of Category,
χ²(4) = 258.67, p < .001, indicating that mean ratings differed reliably across conditions.

The estimated marginal means showed that Combined+ received the highest ratings (estimate = 5.40), whereas all other negative conditions were rated significantly lower. Compared with Combined+, ratings decreased by 1.28 for Biophilic- (p < .001), 0.79 for Ceiling- (p < .001), 2.73 for Combined- (p < .001), and 1.81 for Lighting- (p < .001). Thus, Combined+ produced consistently superior subjective evaluations relative to all negative conditions.

## Posthoc comparisons
```{r model output}
# Estimated marginal means for Category
emm_cat <- emmeans(best_model, ~ Category)

# Residual SD from the model (for standardized effect size d)
resid_sd <- sigma(best_model)

# Pairwise comparisons (Tukey)
pairwise_emm <- pairs(emm_cat, adjust = "tukey")

# Add 95% CI to pairwise results
pairwise_results <- summary(pairwise_emm, infer = TRUE) %>%
  as.data.frame() %>%
    rename(
    Comparison = contrast,
    Difference = estimate,
    Std_Error  = SE,
    DF         = df,
    z_value    = z.ratio,
    P_value    = p.value,
    CI_low     = asymp.LCL,
    CI_high    = asymp.UCL
  ) %>%
  mutate(effect_size = Difference / resid_sd) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

pairwise_results

```

Pairwise comparisons among the negative conditions further revealed that Combined- was rated significantly lower than all other categories (all ps < .001). Additionally, Ceiling- was rated higher than Lighting- (difference = 1.02, p < .001). The difference between Biophilic- and Ceiling- and between Biophilic- and Lighting- did not reach significance (ps > .05), suggesting that these two conditions elicited comparable evaluations.

Taken together, the results indicate a robust effect of Category, with Combined+ performing best, Combined- performing worst, and the remaining negative conditions occupying intermediate positions, with only selective significant differences among them.

```{r echo=FALSE}
emm_df <- as.data.frame(emm_cat)
order = c("Combined+", "Ceiling-", "Biophilic-", "Lighting-", "Combined-")
ggplot(emm_df,
       aes(x = factor(Category, levels = order),
           y = emmean,
           fill = Category)) +
  geom_col(width = 0.7) +
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL),
                position = position_dodge(width = 0.8),
                width = 0.2) +
  labs(
    title = "Model-Predicted Rating by Category",
    x = "Category",
    y = "Predicted Rating"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )
```

