---
title: "Dimensions LMM"
output:
  html_document:
    df_print: paged
---

Analyze each rating dimension separately using LMM.


# Data preparation
## load libraries

```{r load libraries, include=FALSE}
library(tidyverse)
# load the libraries for LMM
library(lme4)
library(lmerTest)
library(car)
library(emmeans)
```

## Read and preprocess data
```{r preprocess, include=FALSE}
# Read the two CSV files
# Set show_col_types = FALSE to keep console clean.
df   <- readr::read_csv("Data_exp1.csv",  show_col_types = FALSE)

# Set Plus as the reference level in Category
df$Category <- relevel(factor(df$Category), ref = "Combined+")
order <- c("Combined+", "Ceiling-", "Biophilic-", "Lighting-", "Combined-")
df$Category <- factor(df$Category, levels = order)

```

# Define functions
## Pairwise comparisons
```{r pairwise}
emm_options(
  lmer.df = "satterthwaite",
  lmerTest.limit = 10000,
  pbkrtest.limit = 0
)



emm_pairwise_table <- function(model, factor_var = "Category", digits = 3) {
  
  # Estimated marginal means
  emm <- emmeans(model, as.formula(paste("~", factor_var)))
  
  # Residual SD (for standardized effect size d)
  resid_sd <- sigma(model)
  
  # Pairwise comparisons (Tukey)
  pw <- pairs(emm, adjust = "tukey")
  
  # Tidy output
  results <- summary(pw, infer = TRUE) %>%
    as.data.frame() %>%
    rename(
      Comparison = contrast,
      Difference = estimate,
      Std_Error  = SE,
      DF         = df,
      t_value    = t.ratio,
      P_value    = p.value,
      CI_low     = lower.CL,
      CI_high    = upper.CL
    ) %>%
    mutate(
      effect_size = Difference / resid_sd
    ) %>%
    mutate(
      across(where(is.numeric), ~ round(.x, digits))
    )
  
  return(results)
}

```

## Plot
```{r plot preditive rating}

plot_predictive_rating <- function(model, factor_var = "Category") {
  
  # Estimated marginal means
  emm <- emmeans(model, as.formula(paste("~", factor_var)))
  emm_df <- as.data.frame(emm)
  order = c("Combined+", "Ceiling-", "Biophilic-", "Lighting-", "Combined-")

  ggplot(emm_df,
       aes(x = factor(Category, levels = order),
           y = emmean,
           fill = Category)) +
  geom_col(width = 0.7) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL),
                position = position_dodge(width = 0.8),
                width = 0.2) +
  labs(
    title = "Model-Predicted Rating by Category",
    x = "Category",
    y = "Predicted Rating"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )
}

```


# LMM

For each rating dimensions, we built the LMM. 

## Hominess
### Model comparisons
```{r}
m1 <- lmer(Hominess ~ Category + (1 | Participant) + (1 | ImageNumber), data=df, REML=FALSE)
m2 <- lmer(Hominess ~ Category + (1 | Participant) + (1 | ImageNumber) +      (0 + Category || Participant), data=df, REML=FALSE)
anova(m1, m2)
```

```{r}
m3 <- lmer(Hominess ~ Category + (1 + Category | Participant) + (1 | ImageNumber), data=df, REML=FALSE)
anova(m2, m3)
anova(m1, m3)
```

### Model output
```{r}
# check the result of best model
best_model_hominess <- m3
summary(best_model_hominess)
Anova(best_model_hominess, type = 'III')
```
### Posthoc comparisons
```{r}
pairwise_hominess <- emm_pairwise_table(best_model_hominess)
pairwise_hominess
```
### Plot
```{r}
plot_predictive_rating(best_model_hominess)
```


## Personalness
### Model comparisons
```{r}
m1 <- lmer(Personalness ~ Category + (1 | Participant) + (1 | ImageNumber), data=df, REML=FALSE)
m2 <- lmer(Personalness ~ Category + (1 | Participant) + (1 | ImageNumber) +      (0 + Category || Participant), data=df, REML=FALSE)
anova(m1, m2)
```

```{r}
m3 <- lmer(Personalness ~ Category + (1 | ImageNumber) +  (1 + Category | Participant), data=df, REML=FALSE)
anova(m3, m2)
anova(m3, m1)
```
### Model output
```{r}
# check the result of best model
best_model_personalness <- m3
summary(best_model_personalness)
Anova(best_model_personalness, type = 'III')
```
### Posthoc comparisons
```{r}
pairwise_personalness <- emm_pairwise_table(best_model_personalness)
pairwise_personalness
```
### Plot
```{r}
plot_predictive_rating(best_model_personalness)
```

## Naturalness
### Model comparisons
```{r}
m1 <- lmer(Naturalness ~ Category + (1 | Participant) + (1 | ImageNumber), data=df, REML=FALSE)
m2 <- lmer(Naturalness ~ Category + (1 | Participant) + (1 | ImageNumber) + (0 + Category || Participant), data=df, REML=FALSE)
anova(m1, m2)
```
```{r}
m3 <- lmer(Naturalness ~ Category + (1 + Category | Participant) + (1 | ImageNumber), data=df, REML=FALSE)
anova(m2, m3)
anova(m1, m3)
```

### Model output
```{r}
# check the result of best model
best_model_naturalness <- m3
summary(best_model_naturalness)
Anova(best_model_naturalness, type = 'III')
```

### Posthoc comparisons
```{r}
pairwise_naturalness <- emm_pairwise_table(best_model_naturalness)
pairwise_naturalness
```

### Plot
```{r}
plot_predictive_rating(best_model_naturalness)
```


## Beauty
### Model comparisons
```{r}
m1 <- lmer(Beauty ~ Category + (1 | Participant) + (1 | ImageNumber), data=df, REML=FALSE)
m2 <- lmer(Beauty ~ Category + (1 | Participant) + (1 | ImageNumber) + (0 + Category || Participant), data=df, REML=FALSE)
anova(m1, m2)
```
```{r}
m3 <- lmer(Beauty ~ Category + (1 + Category | Participant) + (1 | ImageNumber), data=df, REML=FALSE)
anova(m2, m3)
anova(m1, m3)
```
### Model output
```{r}
# check the result of best model
best_model_beauty <- m3
summary(best_model_beauty)
Anova(best_model_beauty, type = 'III')
```

### Posthoc comparisons
```{r}
pairwise_beauty <- emm_pairwise_table(best_model_beauty)
pairwise_beauty
```

### Plot
```{r}
plot_predictive_rating(best_model_beauty)
```


## Approachability
### Model comparisons
```{r}
m1 <- lmer(Approachability ~ Category + (1 | Participant) + (1 | ImageNumber), data=df, REML=FALSE)
m2 <- lmer(Approachability ~ Category + (1 | Participant) + (1 | ImageNumber) + (0 + Category || Participant), data=df, REML=FALSE)
anova(m1, m2)
```
```{r}
m3 <- lmer(Approachability ~ Category + (1 + Category | Participant) + (1 | ImageNumber), data=df, REML=FALSE)
anova(m2, m3)
anova(m1, m3)
```
### Model output
```{r}
# check the result of best model
best_model_approachability <- m3
summary(best_model_approachability)
Anova(best_model_approachability, type = 'III')
```
### Posthoc comparisons
```{r}
pairwise_approachability <- emm_pairwise_table(best_model_approachability)
pairwise_approachability
```

### Plot
```{r}
plot_predictive_rating(best_model_approachability)
```

