---
title: "Greenness_LMM"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

This is a R notebook to analyze the ratings of architectural images using linear mixed models.


# Data preparation

```{r load libraries, include=FALSE}
library(tidyverse)
# load the libraries for LMM
library(lme4)
library(lmerTest)
library(car)
library(emmeans)
```


```{r preprocess, include=FALSE}
# Read the two CSV files
# Set show_col_types = FALSE to keep console clean.
df_exp1   <- readr::read_csv("Data_exp1.csv",  show_col_types = FALSE)
df_imginf <- readr::read_csv("Image_info.csv", show_col_types = FALSE)


# Merge (left join keeps all trials from Data_exp1) ---
# Only bring in the columns needed from Image_info.
df <- df_exp1 %>%
  left_join(df_imginf %>% select(ImageNumber, Greenness), by = "ImageNumber")

# Transform Greenness and set Plus as the reference level in Category
df$Greenness_z <- as.numeric(scale(df$Greenness, center = TRUE, scale = TRUE))
df$Category <- relevel(factor(df$Category), ref = "Combined+")
```

# Modelling
## Find the best model

```{r best model, echo=TRUE, warning=FALSE}
# find the best model
source("select_lmm.R")
best_model <- select_lmm(
  data       = df,
  dv         = "mean_rating",
  fixed_main = c("Category", "Greenness_z"),
  group_subj = "Participant",
  group_item = "ImageNumber",
  verbose    = TRUE
)

```
## Check the output

```{r model output}
# check the result of best model
summary(best_model)
Anova(best_model, type = 'III')
# Save the model object to a file
saveRDS(best_model, file = "best_model.rds")
```

The analysis showed a significant main effect of Category, indicating that mean ratings differed across categories.
In addition, the interaction between Category and Greenness was significant, suggesting that the influence of greenness on ratings varied across categories.
The main effect of Greenness was marginally significant (p = .071), indicating a weak overall trend toward higher ratings with increasing greenness.

## Diagnosis

```{r}
isSingular(best_model, tol = 1e-4)

# Compute the maximum absolute gradient at convergence
oi <- best_model@optinfo
maxgrad <- tryCatch(max(abs(oi$derivs$gradient)), error=function(e) NA)
maxgrad

```
Although the model summary returned a convergence warning (“unable to evaluate scaled gradient; model failed to converge: degenerate Hessian with 1 negative eigenvalue”), we conducted additional diagnostic checks to assess model stability.
The singularity test returned FALSE, indicating that no random-effect variance component was estimated as zero and the model is not overfitted.
Furthermore, the maximum absolute gradient at convergence was very small (max|grad| = 0.0005), well below the recommended threshold of 0.002, suggesting satisfactory numerical convergence.
Taken together, these diagnostics confirm that the model is stable and the results can be interpreted with confidence.

# Interpretation
## Effect of Category
### Post hoc

```{r post hoc, echo=FALSE, message=FALSE, warning=FALSE}
# Estimated marginal means at Greenness_z = 0
emm <- emmeans(best_model, ~ Category, at = list(Greenness_z = 0))

# Pairwise comparisons with Tukey adjustment
pairwise_results <- pairs(emm, adjust = "tukey") %>%
  as.data.frame() %>%
  select(contrast, estimate, SE, df, z.ratio, p.value)

# Clean column names
pairwise_results <- pairwise_results %>%
  rename(
    Comparison = contrast,
    Difference = estimate,
    Std_Error  = SE,
    DF         = df,
    z_value    = z.ratio,
    P_value    = p.value
  )

# Round for nicer output
pairwise_results <- pairwise_results %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

pairwise_results

```

Pairwise comparisons of model-adjusted ratings at the mean level of greenness **(Greenness_z = 0)** were conducted using Tukey-adjusted contrasts.
These comparisons revealed how mean ratings differed across categories after **accounting for greenness**.

Pairwise Tukey-adjusted comparisons revealed that **Combined+** images were rated significantly higher than **Ceiling-** (z = 4.72, p < .001) and **Combined-** (z = 12.77, p < .001) images.
The difference between **Combined+** and **Biophilic-** images did not reach significance (z = 2.37, p = .125), and **Combined+** did not differ from **Lighting-** (p = .961).
Among the categories removing one dimension, Ceiling-, Biophilic-, and Lighting- did not differ significantly from each other (all p > .9), but each was rated higher than **Combined-++ (all p < .001).

Overall, this pattern indicates that **Combined+** images received the highest ratings, **Combined-** the lowest, and the remaining categories did not differ reliably from each other.

### Plot

```{r Mutiple greenness, echo=FALSE}

# Compute model-predicted means at different greenness levels
emm_multi <- emmeans(best_model,
                     ~ Category * Greenness_z,
                     at = list(Greenness_z = c(-2, -1, 0, 1, 2)))

emm_df <- as.data.frame(emm_multi)
order = c("Combined+", "Biophilic-", "Ceiling-", "Lighting-", "Combined-")
ggplot(emm_df,
       aes(x = factor(Greenness_z),
           y = emmean,
           fill = factor(Category, levels = order))) +
  geom_col(position = position_dodge(width = 0.8)) +
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL),
                position = position_dodge(width = 0.8),
                width = 0.2) +
  scale_fill_brewer(palette = "RdYlGn",
                    name = "Category",
                    labels = order) +
  labs(
    title = "Model-Predicted Ratings Across Categories and Greenness Levels",
    x = "Greenness (z-scored)",
    y = "Predicted Rating"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "top",
    legend.title = element_text(face = "bold")
  )
```

The bar plot presents model-predicted mean ratings at five standardized greenness levels (Greenness_z = −2, −1, 0, +1, +2).
Within each greenness level, bars of different colors represent the predicted ratings for each Category, and the error bars indicate 95% confidence intervals.
This visualization allows direct comparison of how different categories were rated under the same greenness condition.

```{r Emmip plot, echo=FALSE}
emmip(best_model, Category ~ Greenness_z,
      at = list(Greenness_z = seq(-2, 2, by = 0.5)),
      CIs = TRUE) +
  labs(
    title = "Interaction between Category and Greenness",
    x = "Greenness (z-scored)",
    y = "Predicted Rating"
  ) +
  theme_minimal()

```

Basically the same content as the previous figure.

## Effect of greenness
### Slopes

To further examine how greenness influenced ratings within each category, we estimated the simple slopes of Greenness_z for every Category.
This analysis provides the rate of change in predicted rating as greenness increases, separately for each category.
The resulting table lists the estimated slopes, standard errors, and 95% confidence intervals, allowing comparison of the strength and direction of greenness effects across categories.

```{r slopes, echo=FALSE, message=FALSE, warning=FALSE}
# Extract the slope of Greenness_z within each Category
trend_res <- emtrends(best_model, ~ Category, var = "Greenness_z")

trend_df <- as.data.frame(trend_res) %>%
  rename(
    Category = Category,
    Slope = Greenness_z.trend,   # Estimated slope for Greenness_z
    Std_Error = SE,              # Standard error of the slope
    DF = df,
    CI_lower = asymp.LCL,        # Lower bound of 95% CI
    CI_upper = asymp.UCL         # Upper bound of 95% CI
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

trend_df

```

### Plot
```{r Slope plot, echo=FALSE}
ggplot(trend_df, aes(x = factor(Category, levels=order), y = Slope)) +
  geom_point(size = 3, color = "darkgreen") +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2, color = "darkgreen") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Effect of Greenness within each Category",
    x = "Category",
    y = "Slope of Greenness (change in rating per 1 SD)"
  ) +
  theme_minimal()
```



These results suggest that greenness does not exert a uniform influence across categories:

In some conditions (e.g., Combined-, possibly Biophilic-), greenness acts as a meaningful predictor of higher ratings.

In others (e.g., Combined+, Ceiling-, Lighting-), the effect of greenness is weak or too uncertain to be conclusive.

The large uncertainty (variance) in the greenness slope under the Lighting condition is primarily driven by participant-level variability, rather than differences between images.The best model did not retain a random slope for greenness by ImageNumber, indicating that greenness effects were relatively consistent across images.

Overall, this pattern supports the idea that the impact of greenness on ratings is context-dependent, rather than consistent across all image types.




