---
title: "Entropy LMM"
output: html_notebook
---

# Data preparation
## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(afex)
library(emmeans)
```

## Read and proprecess data
```{r}
source("load_data.R") 
```



```{r}
# --- Helper Function to Calculate Shannon Entropy ---
# This converts continuous ratings into discrete probability distributions for entropy
calc_shannon <- function(x) {
  # Assuming ratings are integers or can be grouped into 1-7 levels
  p <- table(factor(x, levels = 1:7)) / length(x)
  p <- p[p > 0] # Remove zero probabilities to avoid log(0)
  return(-sum(p * log(p)))
}
```

```{r}
# --- Define the dimensions of interest ---
dims <- c("Hominess", "Naturalness", "Personalness", "Beauty", "Approachability", "mean_rating")

# --- 1. Subject-level Variability (Each row is one Participant per Category) ---
# DF1: SD within each subject
df_subject_sd <- df %>%
  group_by(Participant, Group, Category) %>%
  summarise(across(all_of(dims), list(mean = mean, sd = sd), .names = "{.col}_{.fn}"), .groups = 'drop')

# DF2: Entropy within each subject
df_subject_entropy <- df %>%
  group_by(Participant, Group, Category) %>%
  summarise(across(all_of(dims), list(mean = mean, ent = calc_shannon), .names = "{.col}_{.fn}"), .groups = 'drop')


# --- 2. Item-level Variability (Each row is one ImageID per Group per Category) ---
# DF3: SD across different subjects for the same image
df_item_sd <- df %>%
  group_by(ImageNumber, Group, Category) %>%
  summarise(across(all_of(dims), list(mean = mean, sd = sd), .names = "{.col}_{.fn}"), .groups = 'drop')

# DF4: Entropy across different subjects for the same image
df_item_entropy <- df %>%
  group_by(ImageNumber, Group, Category) %>%
  summarise(across(all_of(dims), list(mean = mean, ent = calc_shannon), .names = "{.col}_{.fn}"), .groups = 'drop')

# --- Final Check ---
cat("Generated DataFrames:\n")
cat("- df_subject_sd: (Subject-level variation using SD)\n")
cat("- df_subject_entropy: (Subject-level variation using Entropy)\n")
cat("- df_item_sd: (Item-level variation using SD)\n")
cat("- df_item_entropy: (Item-level variation using Entropy)\n")

```


# ANOVA

```{r}
# ---  Function: analyze_variability (Dual Compatibility) ---
analyze_variability <- function(data, dv, dv_label) {
  
  # --- 1. Identify Data Level and Structure ---
  if("Participant" %in% names(data)) {
    # Subject-level: Group is Between, Category is Within
    id_var <- "Participant"
    between_var <- "Group"
    within_var <- "Category"
    data_type <- "Subject-level"
  } else if("ImageID" %in% names(data) | "ImageNumber" %in% names(data)) {
    # Item-level: Category is Between, Group is Within
    id_var <- if("ImageID" %in% names(data)) "ImageID" else "ImageNumber"
    between_var <- "Category"
    within_var <- "Group"
    data_type <- "Item-level"
  } else {
    stop("Data must contain either 'Participant' or 'ImageID/ImageNumber'.")
  }

  cat(paste0("\n\n", paste(rep("=", 60), collapse = ""), "\n"))
  cat(paste0("ANALYZING ", data_type, ": ", dv, "\n"))
  cat(paste0(paste(rep("=", 60), collapse = ""), "\n"))

  # --- 2. Run Mixed-Design ANOVA ---
  # aov_ez handles the structure based on between and within arguments
  anova_res <- aov_ez(id = id_var, 
                      dv = dv, 
                      data = data, 
                      between = between_var, 
                      within = within_var)
  
  cat("\n--- ANOVA Results (with Partial Eta Squared) ---\n")
  print(anova_res, es = "pes")
  
  # --- 3. Post-hoc Comparisons & Cohen's d ---
  # We always want to compare Groups within each Category
  emm <- emmeans(anova_res, ~ Group | Category)
  post_hoc_summary <- summary(pairs(emm, reverse = TRUE), infer = c(TRUE, TRUE))
  
  # Extract pooled error (MSE) for effect size calculation
  mse <- anova_res$anova_table$MSE[1]
  pooled_sd <- sqrt(mse)
  
  reporting_table <- post_hoc_summary %>%
    mutate(cohens_d = estimate / pooled_sd) %>%
    select(Category, contrast, estimate, SE, df, lower.CL, upper.CL, t.ratio, p.value, cohens_d)
  
  cat("\n--- Post-hoc Comparisons (Control - ASD) & Cohen's d ---\n")
  print(reporting_table)
  
  # --- 4. Visualization ---
  # Use aes_string or the modern .data[[dv]] for dynamic column names
  plot <- ggplot(data, aes(x = Category, y = .data[[dv]], fill = Group)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.5, width = 0.5, position = position_dodge(0.6)) +
    geom_jitter(aes(color = Group), 
                position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.6), 
                alpha = 0.4, size = 1.5) +
    scale_fill_manual(values = c("ASD" = "#E64B35FF", "Control" = "#4DBBD5FF")) +
    scale_color_manual(values = c("ASD" = "#E64B35FF", "Control" = "#4DBBD5FF")) +
    labs(y = dv_label, x = "Category", title = paste(data_type, "Analysis:", dv)) +
    theme_bw() +
    theme(legend.position = "top", axis.title = element_text(face = "bold"))
  
  print(plot)
  
  return(list(anova = anova_res, post_hoc = reporting_table, plot = plot))
}
```


```{r}
# List of all Entropy columns in the subject_entropy dataframe
target_cols <- c("Hominess_ent", "Naturalness_ent", "Personalness_ent", "Beauty_ent", "Approachability_ent")

# Loop through each and generate reports
for(col in target_cols) {
  analyze_variability(df_subject_entropy, col, paste("Entropy of", gsub("_ent", "", col)))
}
```


```{r}
# List of all SD columns in your subject_sd dataframe
target_cols <- c("mean_rating_sd", "Hominess_sd", "Naturalness_sd", "Personalness_sd", "Beauty_sd", "Approachability_sd")

# Loop through each and generate reports
for(col in target_cols) {
  analyze_variability(df_subject_sd, col, paste("SD of", gsub("_sd", "", col)))
}
```
For all individual dimensions, the interaction effects were significant. Specifically, the SD of ASD group was significantly higher than neurotypical group when viewing Plus images but not Minus Images.

```{r}
# List of all Entropy columns in the item_entropy dataframe
target_cols <- c("Hominess_ent", "Naturalness_ent", "Personalness_ent", "Beauty_ent", "Approachability_ent")

# Loop through each and generate reports
for(col in target_cols) {
  analyze_variability(df_item_entropy, col, paste("Entropy of", gsub("_ent", "", col)))
}
```
In each ASD-CTRL comparison, the entropy of ASD group is higher. 

```{r}
# List of all SD columns in the item_sd dataframe
target_cols <- c("mean_rating_sd", "Hominess_sd", "Naturalness_sd", "Personalness_sd", "Beauty_sd", "Approachability_sd")

# Loop through each and generate reports
for(col in target_cols) {
  analyze_variability(df_item_sd, col, paste("SD of", gsub("_sd", "", col)))
}
```


# Combined plot
## Entropy
```{r}
# --- Function: plot_combined_entropy ---
# Generates a 2x3 grid with 5 subplots (Excluding Mean)
plot_combined_entropy <- function(data) {
  library(tidyverse)
  library(stringr)
  
  # 1. Identify ID variable: Subject-level (Participant) or Item-level (ImageNumber)
  id_var <- if("Participant" %in% names(data)) "Participant" else "ImageNumber"
  
  # 2. Define 5 raw aesthetic dimensions (Excluding Mean)
  dims <- c("Hominess", "Naturalness", "Personalness", "Beauty", "Approachability")
  raw_cols <- paste0(dims, "_ent")
  
  # 3. Pivot to long format
  df_long <- data %>%
    select(any_of(id_var), Group, Category, all_of(raw_cols)) %>%
    pivot_longer(
      cols = all_of(raw_cols),
      names_to = "Dimension",
      values_to = "Value"
    ) %>%
    # Dynamically remove suffix for clean labels
    mutate(Dimension = factor(Dimension, levels = raw_cols, labels = dims))
  
  # 4. Generate the plot
  ggplot(df_long, aes(x = Category, y = Value, fill = Group)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, position = position_dodge(0.6)) +
    geom_jitter(aes(color = Group), 
                position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.6), 
                alpha = 0.3, size = 0.8) +
    facet_wrap(~Dimension, nrow = 2, ncol = 3, scales = "free_y") +
    scale_fill_manual(values = c("ASD" = "#E64B35FF", "Control" = "#4DBBD5FF")) +
    scale_color_manual(values = c("ASD" = "#E64B35FF", "Control" = "#4DBBD5FF")) +
    labs(
      y = "Shannon Entropy (Bits)",
      x = "Stimulus Category",
      title = paste("Rating Uncertainty (Entropy):", 
                   ifelse(id_var == "Participant", "Subject-level", "Item-level"))
    ) +
    theme_bw() +
    theme(
      panel.grid.minor = element_blank(),
      strip.background = element_rect(fill = "grey95"), 
      strip.text = element_text(face = "bold", size = 10),
      legend.position = "top",
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
    )
}
```


## SD

```{r}
# --- Function: plot_combined_sd ---
# Generates a full 2x3 grid with 6 subplots (Including Mean)
plot_combined_sd <- function(data) {
  library(tidyverse)
  library(stringr)
  
  # 1. Identify ID variable
  id_var <- if("Participant" %in% names(data)) "Participant" else "ImageNumber"
  
  # 2. Define all 6 columns (Mean + 5 Dimensions)
  raw_cols <- c("mean_rating_sd", "Hominess_sd", "Naturalness_sd", 
                "Personalness_sd", "Beauty_sd", "Approachability_sd")
  clean_labels <- c("Mean", "Hominess", "Naturalness", "Personalness", "Beauty", "Approachability")
  
  # 3. Pivot to long format
  df_long <- data %>%
    select(any_of(id_var), Group, Category, all_of(raw_cols)) %>%
    pivot_longer(
      cols = all_of(raw_cols),
      names_to = "Dimension",
      values_to = "Value"
    ) %>%
    mutate(Dimension = factor(Dimension, levels = raw_cols, labels = clean_labels))
  
  # 4. Generate the plot
  ggplot(df_long, aes(x = Category, y = Value, fill = Group)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, position = position_dodge(0.6)) +
    geom_jitter(aes(color = Group), 
                position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.6), 
                alpha = 0.3, size = 0.8) +
    facet_wrap(~Dimension, nrow = 2, ncol = 3, scales = "free_y") +
    scale_fill_manual(values = c("ASD" = "#E64B35FF", "Control" = "#4DBBD5FF")) +
    scale_color_manual(values = c("ASD" = "#E64B35FF", "Control" = "#4DBBD5FF")) +
    labs(
      y = "Standard Deviation (SD)",
      x = "Stimulus Category",
      title = paste("Rating Variability (SD):", 
                   ifelse(id_var == "Participant", "Subject-level", "Item-level"))
    ) +
    theme_bw() +
    theme(
      panel.grid.minor = element_blank(),
      strip.background = element_rect(fill = "grey95"), 
      strip.text = element_text(face = "bold", size = 10),
      legend.position = "top",
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
    )
}
```


## Print
```{r}
# 1. Generate Subject-level Plots
p_sub_ent <- plot_combined_entropy(df_subject_entropy)
p_sub_sd  <- plot_combined_sd(df_subject_sd)

# 2. Generate Item-level Plots
p_item_ent <- plot_combined_entropy(df_item_entropy)
p_item_sd  <- plot_combined_sd(df_item_sd)

# 3. Display specific plots
print(p_sub_ent)
print(p_sub_sd)
print(p_item_ent)
print(p_item_sd)
```