---
title: "NeuroArchitecture Exp2 LMM for each dimension ratings"
output: html_notebook
---

# Data preparation
## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(lme4)
library(lmerTest)
library(car)
library(emmeans)
```

## Read data

```{r}
source("load_data.R")
```


# Hominess
## Model selection

```{r}
m1 <- lmer(Hominess ~ Category + Group + (1 | Participant) + (1 | ImageNumber), data=df)
m2 <- lmer(Hominess ~ Category * Group + (1 | Participant) + (1 | ImageNumber), data=df)
anova(m1, m2)
```

```{r}
m3 <- lmer(Hominess ~ Category + Group + (1 + Category | Participant) + (1 | ImageNumber), data=df)
anova(m2, m3)
```

```{r}
m4 <- lmer(Hominess ~ Category * Group + (1 + Category | Participant) + (1 | ImageNumber), data=df)
anova(m3, m4)
```

```{r}
m5 <- lmer(Hominess ~ Category * Group + (1 + Category | Participant) + (1 + Group | ImageNumber), data=df)
anova(m4, m5)
```

## Best model

```{r}
best_model <- m5
summary(best_model)
Anova(best_model)
```
## Post-hoc tests
```{r}
emm_options(pbkrtest.limit = 3700) 
emm_options(lmerTest.limit = 3700)
# This calculates the mean rating for each combination of Group and Category
emm_results <- emmeans(best_model, ~ Group * Category)

# --- 2. Perform Pairwise Comparisons ---
# We split the comparison by Category to see the Group difference in Minus and Plus separately
# 'adjust = "tukey"' handles multiple comparison correction
post_hoc_stats <- pairs(emm_results, by = "Category", reverse = TRUE, adjust = "tukey")

# Display the statistical results (t-ratio, p-values, and estimates)
cat("\n--- Post-hoc Comparisons: Control vs ASD within each Category ---\n")
summary(post_hoc_stats, infer = c(TRUE, TRUE))
```
```{r}
# This focuses on the magnitude of the difference between groups
cat("\n--- Effect Sizes (Cohen's d) ---\n")
eff_size_results <- eff_size(emm_results, sigma = sigma(best_model), edf = df.residual(best_model))

# Extracting only the estimate to keep the output clean as requested
print(eff_size_results)
```

# Naturalness
## Model selection

```{r}
m1 <- lmer(Naturalness ~ Category + Group + (1 | Participant) + (1 | ImageNumber), data=df)
m2 <- lmer(Naturalness ~ Category * Group + (1 | Participant) + (1 | ImageNumber), data=df)
anova(m1, m2)
```



```{r}
m3 <- lmer(Naturalness ~ Category + Group + (1 + Category | Participant) + (1 | ImageNumber), data = df)
anova(m3, m2)
```
```{r}
m4 <- lmer(Naturalness ~ Category * Group + (1 + Category | Participant) + (1 | ImageNumber), data = df)
anova(m3, m4)
```


```{r}
m5 <- lmer(Naturalness ~ Category * Group + (1 + Category | Participant) + (1 + Group | ImageNumber), 
           data = df)
anova(m5, m4)
```
## Best model

```{r}
best_model <- m4
summary(best_model)
Anova(best_model)
```
## Post-hoc tests
```{r}
emm_options(pbkrtest.limit = 3700) 
emm_options(lmerTest.limit = 3700)
# This calculates the mean rating for each combination of Group and Category
emm_results <- emmeans(best_model, ~ Group * Category)

# --- 2. Perform Pairwise Comparisons ---
# We split the comparison by Category to see the Group difference in Minus and Plus separately
# 'adjust = "tukey"' handles multiple comparison correction
post_hoc_stats <- pairs(emm_results, by = "Category", reverse = TRUE, adjust = "tukey")

# Display the statistical results (t-ratio, p-values, and estimates)
cat("\n--- Post-hoc Comparisons: Control vs ASD within each Category ---\n")
summary(post_hoc_stats, infer = c(TRUE, TRUE))
```
```{r}
# This focuses on the magnitude of the difference between groups
cat("\n--- Effect Sizes (Cohen's d) ---\n")
eff_size_results <- eff_size(emm_results, sigma = sigma(best_model), edf = df.residual(best_model))

# Extracting only the estimate to keep the output clean as requested
print(eff_size_results)
```

# Personalness
## Model selection

```{r}
m1 <- lmer(Personalness ~ Category + Group + (1 | Participant) + (1 | ImageNumber), data=df)
m2 <- lmer(Personalness ~ Category * Group + (1 | Participant) + (1 | ImageNumber), data=df)
anova(m1, m2)
```

```{r}
m3 <- lmer(Personalness ~ Category + Group + (1 + Category | Participant) + (1 | ImageNumber), 
           data = df)
anova(m3, m2)
```

```{r}
m4 <- lmer(Personalness ~ Category * Group + (1 + Category | Participant) + (1 | ImageNumber), data = df)
anova(m4, m3)
```

```{r}
m5 <- lmer(Personalness ~ Category * Group + (1 + Category | Participant) + (1 + Group | ImageNumber), data = df)
anova(m5, m4)
```

## Best model

```{r}
best_model <- m5
summary(best_model)
Anova(best_model)
```

## Post-hoc tests
```{r}
emm_options(pbkrtest.limit = 3700) 
emm_options(lmerTest.limit = 3700)
# This calculates the mean rating for each combination of Group and Category
emm_results <- emmeans(best_model, ~ Group * Category)

# --- 2. Perform Pairwise Comparisons ---
# We split the comparison by Category to see the Group difference in Minus and Plus separately
# 'adjust = "tukey"' handles multiple comparison correction
post_hoc_stats <- pairs(emm_results, by = "Category", reverse = TRUE, adjust = "tukey")

# Display the statistical results (t-ratio, p-values, and estimates)
cat("\n--- Post-hoc Comparisons: Control vs ASD within each Category ---\n")
summary(post_hoc_stats, infer = c(TRUE, TRUE))
```
```{r}
# This focuses on the magnitude of the difference between groups
cat("\n--- Effect Sizes (Cohen's d) ---\n")
eff_size_results <- eff_size(emm_results, sigma = sigma(best_model), edf = df.residual(best_model))

# Extracting only the estimate to keep the output clean as requested
print(eff_size_results)
```

# Approachability
## Model selection

```{r}
m1 <- lmer(Approachability ~ Category + Group + (1 | Participant) + (1 | ImageNumber), data=df)
m2 <- lmer(Approachability ~ Category * Group + (1 | Participant) + (1 | ImageNumber), data=df)
anova(m1, m2)
```

```{r}
m3 <- lmer(Approachability ~ Category + Group + (1 + Category | Participant) + (1 | ImageNumber), 
           data = df)
anova(m2, m3)
```
```{r}
m4 <- lmer(Approachability ~ Category * Group + (1 + Category | Participant) + (1 | ImageNumber), 
           data = df)
anova(m3, m4)
```
```{r}
m5 <- lmer(Approachability ~ Category * Group + (1 + Category | Participant) + (1 + Group | ImageNumber), 
           data = df)
anova(m5, m4)
```

## Best model

```{r}
best_model <- m5
summary(best_model)
Anova(best_model)
```
## Post-hoc tests
```{r}
emm_options(pbkrtest.limit = 3700) 
emm_options(lmerTest.limit = 3700)
# This calculates the mean rating for each combination of Group and Category
emm_results <- emmeans(best_model, ~ Group * Category)

# --- 2. Perform Pairwise Comparisons ---
# We split the comparison by Category to see the Group difference in Minus and Plus separately
# 'adjust = "tukey"' handles multiple comparison correction
post_hoc_stats <- pairs(emm_results, by = "Category", reverse = TRUE, adjust = "tukey")

# Display the statistical results (t-ratio, p-values, and estimates)
cat("\n--- Post-hoc Comparisons: Control vs ASD within each Category ---\n")
summary(post_hoc_stats, infer = c(TRUE, TRUE))
```
```{r}
# This focuses on the magnitude of the difference between groups
cat("\n--- Effect Sizes (Cohen's d) ---\n")
eff_size_results <- eff_size(emm_results, sigma = sigma(best_model), edf = df.residual(best_model))

# Extracting only the estimate to keep the output clean as requested
print(eff_size_results)
```

# Beauty
## Model selection

```{r}
m1 <- lmer(Beauty ~ Category + Group + (1 | Participant) + (1 | ImageNumber), data=df)
m2 <- lmer(Beauty ~ Category * Group + (1 | Participant) + (1 | ImageNumber), data=df)
anova(m1, m2)
```

```{r}
m3 <- lmer(Beauty ~ Category + Group + (1 + Category | Participant) + (1 | ImageNumber), data = df)
anova(m3, m2)
```
```{r}
m4 <- lmer(Beauty ~ Category * Group + (1 + Category | Participant) + (1 | ImageNumber), 
           data = df)
anova(m3, m4)
```
```{r}
m5 <- lmer(Beauty ~ Category * Group + (1 + Category | Participant) + (1 + Group | ImageNumber), 
           data = df)
anova(m5, m4)
```
M4 is more satble.

## Best model
```{r}
best_model <- m4
summary(best_model)
Anova(best_model)
```
## Post-hoc tests
```{r}
emm_options(pbkrtest.limit = 3700) 
emm_options(lmerTest.limit = 3700)
# This calculates the mean rating for each combination of Group and Category
emm_results <- emmeans(best_model, ~ Group * Category)

# --- 2. Perform Pairwise Comparisons ---
# We split the comparison by Category to see the Group difference in Minus and Plus separately
# 'adjust = "tukey"' handles multiple comparison correction
post_hoc_stats <- pairs(emm_results, by = "Category", reverse = TRUE, adjust = "tukey")

# Display the statistical results (t-ratio, p-values, and estimates)
cat("\n--- Post-hoc Comparisons: Control vs ASD within each Category ---\n")
summary(post_hoc_stats, infer = c(TRUE, TRUE))
```
```{r}
# This focuses on the magnitude of the difference between groups
cat("\n--- Effect Sizes (Cohen's d) ---\n")
eff_size_results <- eff_size(emm_results, sigma = sigma(best_model), edf = df.residual(best_model))

# Extracting only the estimate to keep the output clean as requested
print(eff_size_results)
```
# Combined plots
Raw rating not estimated marginal means.
```{r}
library(tidyverse)

# --- 1. Aggregate Trial data to Subject-level Means ---
# We calculate the mean for the global score and each individual dimension
subject_means <- df %>%
  group_by(Participant, Group, Category) %>%
  summarise(
    mean_rating = mean(mean_rating, na.rm = TRUE),
    Hominess = mean(Hominess, na.rm = TRUE),
    Naturalness = mean(Naturalness, na.rm = TRUE),
    Personalness = mean(Personalness, na.rm = TRUE),
    Beauty = mean(Beauty, na.rm = TRUE),
    Approachability = mean(Approachability, na.rm = TRUE),
    .groups = 'drop'
  )

# --- 2. Define Columns and Transform to Long Format ---
raw_cols <- c("mean_rating", "Hominess", "Naturalness", 
              "Personalness", "Beauty", "Approachability")

mean_ratings_long <- subject_means %>%
  pivot_longer(
    cols = all_of(raw_cols),
    names_to = "Dimension",
    values_to = "Avg_Rating"
  ) %>%
  mutate(Dimension = factor(Dimension, 
                            levels = raw_cols, 
                            labels = c("Global Mean", "Hominess", "Naturalness", 
                                       "Personalness", "Beauty", "Approachability")))

# --- 3. Create the Combined Plot (2x3 Grid) ---
combined_mean_plot <- ggplot(mean_ratings_long, aes(x = Category, y = Avg_Rating, fill = Group)) +
  # Boxplots
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, position = position_dodge(0.6)) +
  # Individual data points (each point represents one participant's mean)
  geom_jitter(aes(color = Group), 
              position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.6), 
              alpha = 0.4, size = 1) +
  # Use free_y to let each dimension show its own range clearly
  facet_wrap(~Dimension, nrow = 2, ncol = 3, scales = "free_y") +
  # Scientific color palette
  scale_fill_manual(values = c("ASD" = "#E64B35FF", "Control" = "#4DBBD5FF")) +
  scale_color_manual(values = c("ASD" = "#E64B35FF", "Control" = "#4DBBD5FF")) +
  labs(
    y = "Average Rating (1-7)",
    x = "Stimulus Category",
    title = "Subject-level Mean Ratings Across Dimensions"
  ) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "grey95"), 
    strip.text = element_text(face = "bold"),
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

# --- 4. Print Plot ---
print(combined_mean_plot)
```

